Revealing transient strain in geodetic data with Gaussian process regression

### 2017-10-29 RevisedSubmission/hines_hetland_2017_gji.tex
did more shit

By choosing a FOGM process for $z_i$, we have introduced two
parameters that need to be constrained, $\alpha$ and $\beta$. We
constrain these parameters, which we collectively refer to as
$\mitbf{\theta}$, with the Restricted Maximum Likelihood (REML) method
\citep{Harville1974}. The REML method is conceptually similar to the
Maximum Likelihood Estimation (MLE) method from \citet{Langbein1997},
in that we attempt to pick $\mitbf{\theta}$ to maximize the
probability of drawing the observed data, $\mitbf{d}_i^*$, from the
random vector $\mitbf{d}_i$. However, unlike the MLE method, the REML
method produces unbiased estimates of $\mitbf{\theta}$ \citep[sec.
2.6]{Cressie1992}. We use the REML method to estimate $\mitbf{\theta}$
at 38 continuous GNSS stations in the Pacific Northwest that are east
of 121$^\circ$W. These stations are sufficiently far from the
subduction zone that they are unlikely to contain transient signal
associated with SSEs, allowing us to ignore displacements from $u_i$
in $\mitbf{d}_i$. We assume the noise at these inland stations is
representative of the noise at all the stations considered in this
study, which is probably a poor assumption since the inland stations
are subject to distinctly different climatic conditions. Nonetheless,
we find $\mitbf{\theta}$ for each of the inland stations and for each
displacement component by maximizing the REML likelihood function

\begin{equation}\label{eq:REML}
\mathcal{L}(\mitbf{\theta}) = \left(\frac{\left|\G^T\G\right|}{(2\pi)^{n-6m} 
                                    \left| \mitbf{C}_{\eta_i}(\mitbf{\theta}) \right| 
                                    \left| \G^T\mitbf{C}_{\eta_i}(\mitbf{\theta})^{-1}\G \right|}\right)^{\frac{1}{2}} 
                              e^{-\tfrac{1}{2}\data_i^{*T}\mitbf{K}(\mitbf{\theta})\data_i^*}
\end{equation}
with respect to $\mitbf{\theta}$, where
\begin{equation}
\mitbf{K}(\mitbf{\theta}) = \mitbf{C}_{\eta_i}(\mitbf{\theta})^{-1} - 
                            \mitbf{C}_{\eta_i}(\mitbf{\theta})^{-1}\G
                            \left(\G^T\mitbf{C}_{\eta_i}(\mitbf{\theta})^{-1}\G\right)^{-1}
                            \G^T\mitbf{C}_{\eta_i}(\mitbf{\theta})^{-1}.
\end{equation}

@@

We next establish our prior model for transient displacements.
Specifically, we discuss our choice for the functions $X_i$ and $T_i$.
For $X_i$, we use the squared exponential (SE) function,
\begin{equation}\label{eq:SE}
X_i(\pos,\pos') = \exp\left(\frac{-||\pos - \pos'||_2^2}{2 \ell^2}\right).
\end{equation}
As we have mentioned, the SE function is commonly used for kriging
\citep[e.g,][]{Cressie1992} and Gaussian process regression
\citep[e.g.,][]{Rasmussen2006}. In terms of geodetic applications,
\citet{Kato1998} and \cite{El-Fiky1999} demonstrated that the SE
function accurately describes the spatial covariance of secular GNSS
derived velocities in Japan.

@@

We consider three potential choices for $T_i$. The first option is the
one-dimensional SE function,
\begin{equation}\label{eq:TimeSE}
T_i(t,t') = \phi^2\exp\left(\frac{-|t - t'|^2}{2\tau^2}\right).
\end{equation}
Note that $T_i$ includes the parameter $\phi$, which serves to scale
the covariance function $C_{u_i}$. The second option is the Wendland
class of covariance functions \citep{Wendland2005}. Wendland
covariance functions have compact support and hence their
corresponding covariance matrices are sparse. In our analysis, we
exploit this sparsity with the CHOLMOD software package
\citep{Chen2008}. Wendland functions are positive definite in
$\mathbb{R}^d$, and they can be differentiated $k$ times. The form of
the covariance function depends on the choice of $d$ and $k$. We use
$d=1$ since we are describing the temporal covariance of $u_i$. We use
$k=2$, giving samples of $u_i$ continuous velocities and
accelerations. The corresponding Wendland covariance function is
\begin{equation}\label{eq:Wendland} 
T_i(t,t') = \phi^2\left(1 - \frac{|t - t'|}{\tau}\right)^5_+ 
            \left(\frac{8|t - t'|^2}{\tau^2} + \frac{5|t - t'|}{\tau} + 1\right), 
\end{equation} 
where $(t)_+ = \mathrm{min}(0,t)$.
The third option for $T_i$ is the covariance function for integrated
Brownian motion (IBM). IBM is a Gaussian process with zero mean and
its covariance function can be found by integrating the covariance
function for Brownian motion,
\begin{align}\label{eq:IBM}
T_i(t,t') &= \int_0^t \int_0^{t'} \phi^2 \min(s,s') \,ds'\,ds \\
          &= \frac{\phi^2}{2}\min(t,t')^2 \left(\max(t,t') - \frac{1}{3}\min(t,t')\right), \ \ \ t,t' \geq 0.
\end{align}
IBM has been used in the context of Kalman filtering as a
non-parametric model for the time dependence of geophysical signals
\citep[e.g.,][]{Segall1997, McGuire2003, Ohtani2010, Hines2016a}.
Similar to Brownian motion, IBM has a reference time, $t=0$, at which
the process begins. For some geophysical signals, it is appropriate to
have this reference time. For example, if we are trying to identify
postseismic deformation then $t$ should be zero at the time of the
earthquake.  However, if we are interesting in detecting transient
events, where there is no known start time, then IBM may not be an
appropriate prior. Instead, one should prefer an isotropic choice for
$T_i$, meaning that it is invariant to time shifts and can be written
in terms of $|t - t'|$. In the following analysis, we make the choice
that $t$ is zero on the first epoch of $\data_i^*$. Using an earlier
reference time does not change the results discussed in this section.

@@

% SHOW REML EQUATION IN HERE

We next use the REML method to determine appropriate parameters for
$X_i$ and each of the three choices for $T_i$. we divide the dataset
into seven subsets which are four months long and each centered on the
time of a SSE. The times of the seven SSEs are determined with tremor
records from \cite{Wech2010}. We find the optimal parameters for $T_i$
and $X_i$ for each subset of data. We choose to make each data subsets
four months long because it is long enough to encompass a SSE in the
Pacific Northwest, while it is short enough to still be
computationally tractable. However, four months is too short to
resolve the sinusoids in $\data_i$, and they are omitted from
$\data_i$ in this REML analysis for Pacific Northwest SSEs.

% Results for our parameters

The estimated hyperparameters for $u$ are summarized in Table 1. Based
on the interquartile ranges, the estimated hyperparameters for the SE
and Wendland covariance functions do not vary significantly between
SSEs. This suggests that the median of estimated hyperparameters
should be an appropriate prior model for all Pacific Northwest SSEs.
For the IBM model, there are several anomalously large values for
$\ell$ and $\phi$, which results in large interquartile ranges.

\begin{table}\label{tab:Parameters}
\caption{
Optimal hyperparameters for the prior on transient displacements
determined with the REML method. The temporal covariance function is
indicated by the ``$T$'' column. The SE, IBM, and Wendland covariance
functions are defined in eqs. (\ref{eq:TimeSE}), (\ref{eq:IBM}), and
(\ref{eq:Wendland}), respectively. The spatial covariance function,
$X$, is the squared exponential (eq. \ref{eq:SE}) in all cases. The
hyperparameters are estimated for each of the seven SSEs considered in
this study, and the tabulated values indicate the median and
interquartile ranges of estimates. The ``diff $\log$(REML)'' column
compares the log REML likelihood to the log REML likelihood when using
the SE covariance function for $T$. Negative values indicate that
observations are more consistent with the SE covariance function.
} 
\begin{tabular} {l l l l l l}
$T$ & direction & $\ell$  & $\phi$   & $\tau$  & diff. $\log$(REML) \\ \hline
SE & east   & 92 $\pm$ 25 km  & 0.62 $\pm$ 0.11 mm  & 0.026 $\pm$ 0.011 yr  &  - \\
SE & north  & 91 $\pm$ 53 km  & 0.43 $\pm$ 0.05 mm  & 0.030 $\pm$ 0.017 yr  &  - \\
Wendland & east   & 95 $\pm$ 30 km  & 0.66 $\pm$ 0.15 mm  & 0.093 $\pm$ 0.044 yr &  0.78 $\pm$ 0.87 \\
Wendland & north  & 92 $\pm$ 57 km  & 0.46 $\pm$ 0.10 mm  & 0.116 $\pm$ 0.057 yr &  0.08 $\pm$ 0.58 \\
IBM & east   & 110 $\pm$ 130 km & 290 $\pm$ 420 mm/yr$^{1.5}$  & -          & -16.4 $\pm$ 7.8 \\
IBM & north  & 150 $\pm$ 560 km & 110 $\pm$ 250 mm/yr$^{1.5}$ & -           & -10.1 $\pm$ 2.3 \\
\end{tabular}
\end{table}

@@

To further explore which covariance function for $T$ best describes
SSEs, we compare the observations to the predicted displacements for
each covariance function. We consider the data prediction vector to be
$\hat{\data} = \left(u(\points) + \G\mitbf{a}\right)|\data_*$.
Following a similar procedure as in Section \ref{sec:Method}, it can
be shown that $\hat{\data}$ is normally distributed with mean
\begin{equation}\label{eq:DataPredMean}
\mitbf{\mu}_{\hat{d}} = \left[\begin{array}{cc}
                           C_u(\points,\points) & \G \\
                           \end{array}\right]
                     \left[\begin{array}{cc}
                           \mitbf{\Sigma} & \G \\
                           \G^T  & \zeros \\
                           \end{array}\right]^{-1}
                     \left[\begin{array}{c}
                           \data_* \\
                           \zeros \\
                           \end{array}\right]
\end{equation}  
and covariance
\begin{equation}\label{eq:DataPredCov}
\mitbf{C}_{\hat{\data}} = C_u(\points,\points) - 
                        \left[\begin{array}{cc}
                              C_u(\points,\points) & \G \\
                              \end{array}\right]
                        \left[\begin{array}{cc}
                              \mitbf{\Sigma} & \G \\
                              \G^T  & \zeros \\
                              \end{array}\right]^{-1}
                        \left[\begin{array}{c}
                              C_u(\points,\points) \\
                              \G^T \\
                              \end{array}\right].
\end{equation}
We compute $\hat{\data}$ using SE, Wendland, and IBM covariance
functions for $T$ and the median hyperparameters from Table 1. Figure
\ref{fig:Fit} compares the easting component of $\data_*$ to
$\hat{\data}$ during the winter 2015-2016 SSE at the three stations
that record the strongest SSE signal, ALBH, P436, and SC02. Based on
Figure \ref{fig:Fit}, $\hat{\data}$ is insensitive to the choice of
covariance function for $T$. The predicted displacements for the IBM
covariance function contain slightly more high frequency, and perhaps
spurious, features. When comparing $\hat{\data}$ to $\data_*$ at
station ALBH and SC02, it appears that $\hat{\data}$ slightly
underestimates the rate of deformation during the SSE, regardless of
the chosen covariance function. This over-smoothing could indicate
that the chosen time-scale hyperparameter, $\tau$, is too large. The
predicted displacements at the remaining stations, which record more
subtle SSE deformation, seem to faithfully describe $\data_*$.

For our estimates of transient strain discussed in the next section,
we ultimately settle on the Wendland covariance function for $T$ and
use the median values from Table 1 for the hyperparameters. We choose
the Wendland covariance function over the SE covariance function
because of its computational advantages.

>>>>

By choosing a FOGM process for $z_i$, we have introduced two
parameters that need to be constrained, $\alpha$ and $\beta$. We
constrain these parameters, which we collectively refer to as
$\mitbf{\theta}$, with the Restricted Maximum Likelihood (REML) method
\citep{Harville1974}. The REML method is conceptually similar to the
Maximum Likelihood Estimation (MLE) method from \citet{Langbein1997},
where we pick $\mitbf{\theta}$ to maximize the probability of drawing
the observed data, $\mitbf{d}_i^*$, from the random vector
$\mitbf{d}_i$. However, unlike the MLE method, the REML method
produces unbiased estimates of $\mitbf{\theta}$ \citep[sec.
2.6]{Cressie1992}. We use the REML method to estimate $\mitbf{\theta}$
at 38 continuous GNSS stations in the Pacific Northwest that are east
of 121$^\circ$W. These stations are sufficiently far from the
subduction zone that they are unlikely to contain transient signal
associated with SSEs, allowing us to ignore displacements from $u_i$
in $\mitbf{d}_i$. We assume the noise at these inland stations is
representative of the noise at all the stations considered in this
study, which is probably a poor assumption since the inland stations
are subject to distinctly different climatic conditions. Nonetheless,
we find $\mitbf{\theta}$ for each of the inland stations and for each
displacement component by maximizing the REML likelihood function
\begin{equation}\label{eq:REMLNoise}
\mathcal{L}(\mitbf{\theta}) = \left(\frac{\left|\G^T\G\right|}{(2\pi)^{n-6m} 
                                    \left| \mitbf{C}_{\eta_i}(\mitbf{\theta}) \right| 
                                    \left| \G^T\mitbf{C}_{\eta_i}(\mitbf{\theta})^{-1}\G \right|}\right)^{\frac{1}{2}} 
                              e^{-\tfrac{1}{2}\data_i^{*T}\mitbf{K}(\mitbf{\theta})\data_i^*}
\end{equation}
with respect to $\mitbf{\theta}$, where
\begin{equation}
\mitbf{K}(\mitbf{\theta}) = \mitbf{C}_{\eta_i}(\mitbf{\theta})^{-1} - 
                            \mitbf{C}_{\eta_i}(\mitbf{\theta})^{-1}\G
                            \left(\G^T\mitbf{C}_{\eta_i}(\mitbf{\theta})^{-1}\G\right)^{-1}
                            \G^T\mitbf{C}_{\eta_i}(\mitbf{\theta})^{-1}.
\end{equation}
In the above equation, $\data^*_i$ consists of the data for a single station.

@@

We next establish our prior model for transient displacements.
Specifically, we discuss our choice for the functions $X_i$ and $T_i$.
For $X_i$, we use the squared exponential (SE) covariance function,
\begin{equation}\label{eq:SE}
X_i(\pos,\pos') = \exp\left(\frac{-||\pos - \pos'||_2^2}{2 \ell^2}\right).
\end{equation}
The SE covariance function is commonly used for kriging
\citep[e.g,][]{Cressie1992} and Gaussian process regression
\citep[e.g.,][]{Rasmussen2006}. In terms of geodetic applications,
\citet{Kato1998} and \cite{El-Fiky1999} demonstrated that the SE
covariance function accurately describes the spatial covariance of
secular GNSS derived velocities in Japan.

@@

We consider three potential choices for $T_i$. The first option is the
one-dimensional SE covariance function,
\begin{equation}\label{eq:TimeSE}
T_i(t,t') = \phi^2\exp\left(\frac{-|t - t'|^2}{2\tau^2}\right).
\end{equation}
Note that $T_i$ includes the parameter $\phi$, which serves to scale
the covariance function $C_{u_i}$. The second option is a member of
the Wendland class of covariance functions \citep{Wendland2005}.
Wendland covariance functions have compact support and hence their
corresponding covariance matrices are sparse. In our analysis, we
exploit this sparsity with the CHOLMOD software package
\citep{Chen2008}. Wendland covariance functions can be constructed
such that they are positive definite in $\mathbb{R}^d$ and their
corresponding Gaussian process can be differentiated $k$ times. We use
$d=1$, because we are describing the temporal covariance of $u_i$, and
we use $k=2$, giving samples of $u_i$ continuous velocities and
accelerations. The corresponding Wendland covariance function is
\begin{equation}\label{eq:Wendland} 
T_i(t,t') = \phi^2\left(1 - \frac{|t - t'|}{\tau}\right)^5_+ 
            \left(\frac{8|t - t'|^2}{\tau^2} + \frac{5|t - t'|}{\tau} + 1\right), 
\end{equation} 
where $(t)_+ = \mathrm{max}(0,t)$. The third option for $T_i$ is the
covariance function for integrated Brownian motion (IBM). IBM is a
Gaussian process with zero mean and its covariance function can be
found by integrating the covariance function for Brownian motion,
\begin{align}\label{eq:IBM}
T_i(t,t') &= \int_0^t \int_0^{t'} \phi^2 \min(s,s') \,ds'\,ds \\
          &= \frac{\phi^2}{2}\min(t,t')^2 \left(\max(t,t') - \frac{1}{3}\min(t,t')\right), \ \ \ t,t' \geq 0.
\end{align}
IBM has been used in the context of Kalman filtering as a
non-parametric model for the time dependence of geophysical signals
\citep[e.g.,][]{Segall1997, McGuire2003, Ohtani2010, Hines2016a}.
Similar to Brownian motion, IBM has a reference time, $t=0$, at which
the process begins. For some geophysical signals, it is appropriate to
have this reference time. For example, if we are trying to identify
postseismic deformation then $t$ should be zero at the time of the
earthquake.  However, if we are interesting in detecting transient
events, where there is no known start time, then IBM may not be an
appropriate prior model. Instead, one may prefer to use the SE or
Wendland covariance functions because they are stationary. In the
following analysis, we make the choice that $t$ is zero on the first
epoch of $\data_i^*$. Using an earlier reference time does not change
the results discussed in this section.

@@

We use the REML method to determine the appropriate values for the
parameters in $X_i$ and $T_i$, $\mitbf{\theta} = \{\ell, \phi, \tau
\}$. we divide the GNSS data into seven subsets that are four months
long and each centered on the time of a SSE. The times of the seven
SSEs are determined with tremor records from \cite{Wech2010}. We find
the optimal $\mitbf{\theta}$ for each subset of data, for each
displacement component, and for each choice of $T_i$. We choose to
make each data subsets four months long because it is long enough to
encompass a SSE in the Pacific Northwest, while it is still
computationally tractable to estimate $\mitbf{\theta}$. However, four
months is too short to resolve the sinusoids in $\data_i$, and they
are omitted from $\data_i$ in this REML analysis. The REML likelihood
function that we are maximizing with respect to $\mitbf{\theta}$ is
now 
\begin{equation}\label{eq:REMLPrior} 
\mathcal{L}(\mitbf{\theta}) =
\left(\frac{\left| \G^T \G \right|}{(2\pi)^{n-6m}
                                    \left| \mitbf{\Sigma}(\mitbf{\theta}) \right| 
                                    \left| \G^T \mitbf{\Sigma}(\mitbf{\theta})^{-1} \G \right|}\right)^{\frac{1}{2}} 
                              e^{-\tfrac{1}{2} \data_i^{*T} \mitbf{K}(\mitbf{\theta}) \data_i^*},
\end{equation}
where
\begin{equation}
\mitbf{K}(\mitbf{\theta}) = \mitbf{\Sigma}(\mitbf{\theta})^{-1} - 
                            \mitbf{\Sigma}(\mitbf{\theta})^{-1} \G
                            \left( \G^T \mitbf{\Sigma}(\mitbf{\theta})^{-1} \G \right)^{-1}
                            \G^T \mitbf{\Sigma}(\mitbf{\theta})^{-1}
\end{equation}
and $\mitbf{\Sigma}(\mitbf{\theta}) = \mitbf{C}_{\eta_i} +
C_{u_i}(\points,\points;\mitbf{\theta})$. In the above equation,
$\data^*_i$ consists of a single subset of data. The estimated
parameters are summarized in Table 1. Based on the interquartile
ranges, the estimated parameters for the SE and Wendland covariance
functions do not vary significantly between SSEs. This suggests that
the median of estimated parameters should be an appropriate prior
model for all Pacific Northwest SSEs. For the IBM model, there are
several anomalously large values for $\ell$ and $\phi$, which is why
the interquartile ranges are large.

\begin{table}\label{tab:Parameters}
\caption{
Optimal parameters for the prior on transient displacements determined
with the REML method. The temporal covariance function is indicated by
the ``$T$'' column. The SE, IBM, and Wendland covariance functions are
defined in eqs. (\ref{eq:TimeSE}), (\ref{eq:IBM}), and
(\ref{eq:Wendland}), respectively. The spatial covariance function,
$X$, is the squared exponential (eq. \ref{eq:SE}) in all cases. The
parameters are estimated for each of the seven SSEs considered in
this study, and the tabulated values indicate the median and
interquartile ranges of estimates. The ``diff $\log$(REML)'' column
compares the log REML likelihood to the log REML likelihood when using
the SE covariance function for $T$. Negative values indicate that
observations are more consistent with the SE covariance function.
} 
\begin{tabular} {l l l l l l}
$T$ & direction & $\ell$  & $\phi$   & $\tau$  & diff. $\log$(REML) \\ \hline
SE & east   & 92 $\pm$ 25 km  & 0.62 $\pm$ 0.11 mm  & 0.026 $\pm$ 0.011 yr  &  - \\
SE & north  & 91 $\pm$ 53 km  & 0.43 $\pm$ 0.05 mm  & 0.030 $\pm$ 0.017 yr  &  - \\
Wendland & east   & 95 $\pm$ 30 km  & 0.66 $\pm$ 0.15 mm  & 0.093 $\pm$ 0.044 yr &  0.78 $\pm$ 0.87 \\
Wendland & north  & 92 $\pm$ 57 km  & 0.46 $\pm$ 0.10 mm  & 0.116 $\pm$ 0.057 yr &  0.08 $\pm$ 0.58 \\
IBM & east   & 110 $\pm$ 130 km & 290 $\pm$ 420 mm/yr$^{1.5}$  & -          & -16.4 $\pm$ 7.8 \\
IBM & north  & 150 $\pm$ 560 km & 110 $\pm$ 250 mm/yr$^{1.5}$ & -           & -10.1 $\pm$ 2.3 \\
\end{tabular}
\end{table}

@@

To further explore which covariance function for $T$ best describes
SSEs, we compare the observations to the predicted displacements for
each covariance function. We consider the data prediction vector to be
$\hat{\data} = \left(u(\points) + \G\mitbf{a}\right)|\data_*$.
Following a similar procedure as in Section \ref{sec:Method}, it can
be shown that $\hat{\data}$ is normally distributed with mean
\begin{equation}\label{eq:DataPredMean}
\mitbf{\mu}_{\hat{d}} = \left[\begin{array}{cc}
                           C_u(\points,\points) & \G \\
                           \end{array}\right]
                     \left[\begin{array}{cc}
                           \mitbf{\Sigma} & \G \\
                           \G^T  & \zeros \\
                           \end{array}\right]^{-1}
                     \left[\begin{array}{c}
                           \data_* \\
                           \zeros \\
                           \end{array}\right]
\end{equation}  
and covariance
\begin{equation}\label{eq:DataPredCov}
\mitbf{C}_{\hat{\data}} = C_u(\points,\points) - 
                        \left[\begin{array}{cc}
                              C_u(\points,\points) & \G \\
                              \end{array}\right]
                        \left[\begin{array}{cc}
                              \mitbf{\Sigma} & \G \\
                              \G^T  & \zeros \\
                              \end{array}\right]^{-1}
                        \left[\begin{array}{c}
                              C_u(\points,\points) \\
                              \G^T \\
                              \end{array}\right].
\end{equation}
We compute $\hat{\data}$ using SE, Wendland, and IBM covariance
functions for $T$ and the median parameters from Table 1. Figure
\ref{fig:Fit} compares the easting component of $\data_*$ to
$\hat{\data}$ during the winter 2015-2016 SSE at the three stations
that record the strongest SSE signal, ALBH, P436, and SC02. Based on
Figure \ref{fig:Fit}, $\hat{\data}$ is insensitive to the choice of
covariance function for $T$. The predicted displacements for the IBM
covariance function contain slightly more high frequency, and perhaps
spurious, features. When comparing $\hat{\data}$ to $\data_*$ at
station ALBH and SC02, it appears that $\hat{\data}$ slightly
underestimates the rate of deformation during the SSE, regardless of
the chosen covariance function. This over-smoothing could indicate
that the chosen time-scale parameter, $\tau$, is too large. The
predicted displacements at the remaining stations, which record more
subtle SSE deformation, seem to faithfully describe $\data_*$.

For our estimates of transient strain discussed in the next section,
we ultimately settle on the Wendland covariance function for $T$ and
use the median values from Table 1 for the parameters. We choose
the Wendland covariance function over the SE covariance function
because of its computational advantages.
